How to Avoid OOTA Without Really Trying

### 2024-10-03 paper/ootavacuous.tex
Improve the examples of invented loads
Explain the roles of r1a and r1b in 4.3 and explain why these results
lead us to prohibit inventing atomic loads.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

However, an especially devious compiler might transform the source into
the following form before translating it to machine code:
\begin{quote}
\begin{verbatim}
 1. int r1;
 2. int r1a = (x != 0);
 3. int r1b = (x != 0);
 4. int r2 = (y != 0);
 5. if (r1a != r1b) {
 6.     r1 = r2;
 7.     z = 1;
 8. } else {
 9.     r1 = r1b;
10.     z = (r1 == r2);
11. }
\end{verbatim}
\end{quote}
The idea is that \co{r1a}, \co{r1b}, and \co{r2} can each be only zero or one,
so if \co{r1a} and \co{r1b} differ then one of them must be equal to \co{r2}.
In executions where this happens---because another thread writes to \co{x}
between the two loads---the implementation can choose at runtime to use
for \co{r1} whichever value agrees with \co{r2}, as shown on line~6.
Then the value stored to \co{z} on line~7 will simply be one,
with no dependence on the value loaded from \co{y}.

Therefore we should rule out transformations like these by insisting
the compiler not invent (or duplicate) atomic loads.
In fact, we will require that atomic accesses be treated as
``quasi volatile'', in that the compiler is allowed to omit,
merge, or reorder them but not invent them.

>>>>

However, an especially devious compiler might transform the source into
the following form before translating it to machine code:
\begin{quote}
\begin{verbatim}
 1. int r1;
 2. int r1a = (x != 0);
 3. int r1b = (x != 0);
 4. int r2 = (y != 0);
 5. if (r1a != r1b) {
 6.     r1 = r2;
 7.     z = 1;
 8. } else {
 9.     r1 = r1a;
10.     z = (r1 == r2);
11. }
\end{verbatim}
\end{quote}
Here \co{r1a} represents the original load from \co{x} while the \co{r1b}
load is invented by the compiler.

The idea is that \co{r1a}, \co{r1b}, and \co{r2} can each be only zero or one,
so if \co{r1a} and \co{r1b} differ then one of them must be equal to \co{r2}.
In executions where this happens---because another thread writes to \co{x}
between the two loads---the implementation can choose at runtime to use
for \co{r1} whichever value agrees with \co{r2}, as shown on line~6.
Then the value stored to \co{z} on line~7 will simply be one,
with no dependence on the value loaded from \co{y}.

Because invented loads can cause errors and break dependencies,
as we have just seen, we should insist that
the compiler not invent (or duplicate) atomic loads.
In fact, we will require that atomic accesses be treated as
``quasi volatile'', in that the compiler is allowed to omit,
merge, or reorder them but not invent them.
